Performance portability of an intermediate-complexity atmospheric research model in coarray Fortran

### 2017-08-16 body.tex
Add Sec 1.2 Objectives; edit 2.1 Physics &Numerics

\section{Introduction}
\subsection{Motivation and Background}
Since the Fortran 2008 standard was published in 2010~\cite{iso2010information}, a handful of studies in the open literature
have provided encouraging assessments of Fortran's intrinsic parallel programming model, \gls{caf}, in research
applications running at scale~\cite{preissl2011multithreaded,garain2015comparing,mozdzynski2015partitioned}.   One of the
most attractive features of \gls{caf} is the ability to express parallel algorithms once within the confines of a single
language standard without embedding compiler directives or referencing communication layers that are not part of the
language standard (e.g., no \gls{mpi} or OpenMP).  In theory, this should leave the choice of communication layers
as a link-time decision.  Ideally, an application developer writes a standard, Fortran-only application, compiles it with any
standard-conforming compiler, and possibly reserves the option to incorporate any one (or more) of several communication layers
at link-time.

@@

\subsection{Objectives}

\section{Methodology}
\subsection{Numerical algorithms}
The core numerics of \gls{icar} implemented here are the Thompson Eidhammer microphysics parameterization~\cite{Thompson:2014cw} and the first order MP-DATA advection algorithm~\cite{Smolarkiewicz:1998il}.
The microphysics parameterization requires 11 primary input variables (e.g.\ air pressure, water vapor, cloud water), 9 of which are modified by the parameterization and advected throughout the domain requiring communication between processes.

The test case implemented here uses a classical idealized hill case.
An idealized 1000 m high mountain is generated in the middle of the domain defined by a sine curve.
The air is initialized to near 100\% relative humidity, and a background wind of approximately 10 m s$^{-1}$ is specified.

>>>>

\section{Introduction}
\subsection{Motivation and Background}
Since the Fortran 2008 standard was published in 2010~\cite{iso2010information}, a handful of published studies
have provided encouraging assessments of Fortran's intrinsic parallel programming model, \gls{caf}, in research
applications running at scale~\cite{preissl2011multithreaded,garain2015comparing,mozdzynski2015partitioned}.   One of the
most attractive features of \gls{caf} is the ability to express parallel algorithms once within the confines of a single
language standard without embedding compiler directives or referencing communication layers that are not part of the
language standard (e.g., no \gls{mpi} or OpenMP).  In theory, this should leave the choice of communication layers
as a link-time decision.  Ideally, an application developer writes a standard, Fortran-only application, compiles it with any
standard-conforming compiler, and possibly reserves the option to incorporate any one (or more) of several communication layers
at link-time.

@@

\subsection{Objectives}
The goals of this work are to investigate and evaluate several software and hardware alternatives for compiling and linking the
same \gls{mini-app} \gls{caf} source code using the following technologies:
\begin{itemize}
  \item \gls{mpi}~\cite{mpiforum2016mpi} and OpenSHMEM~\cite{chapman2010introducing} communication layers,
  \item \gls{gcc} and Cray compilers, and
  \item Multi- and many-core processors.
\end{itemize}
Additionally we show performance differences between two one-sided data access patterns:
\begin{enumerate}
  \item Gets: an image that needs data proactively gets the data from memory managed by a providing
        image without the providing image's active involvement, and
  \item Puts: an image that is able to provide data proactively puts the data into memory managed by
        a receiving image without the receiving image's active involvement.
\end{enumerate}
This study is a work-in-progress that will guide future decisions around the choice of compiler, communication
layer, platform, and data access patterns.

\section{Methodology}
\subsection{Physics and numerics}
The coarray-\gls{icar} implemention employed here uses the Thompson Eidhammer microphysics parameterization~\cite{Thompson:2014cw} and the first-order MP-DATA advection algorithm~\cite{Smolarkiewicz:1998il}.
The microphysics parameterization requires 11 primary input variables (e.g.\ air pressure, water vapor, cloud water), 9 of which are modified by the parameterization and advected throughout the domain requiring communication between processes.

The test case implemented here uses a classical, idealized hill.
An idealized 1000 m high mountain is generated in the middle of the domain defined by a sine curve.
The air is initialized to near 100\% relative humidity, and a background wind of approximately 10 m s$^{-1}$ is specified.
